\documentclass{article}
\usepackage[hypertex]{hyperref}
\def\PAL{\mbox{C{\texttt{-{}-}}}}
% l2h substitution PAL C--
\title{Release notes for the Quick~{\PAL} compiler}
\author{The Quick~{\PAL} Team}

\begin{document}

\maketitle

\tableofcontents

\clearpage

\section{Unpleasant surprises}

\subsection{Floating-point computation}

\begin{itemize}
\item
Floating-point literals aren't implemented.
\item
The manual presents two styles of doing floating-point comparisons,
but only one style (\texttt{feq}, \texttt{flt}, \ldots) is actually
implemented---you can't usefully use \texttt{fcmp}.
\item
The only rounding modes that are supported are the hardware rounding
modes of the target machine.
These need to be declared as a special global register variable, e.g.,
\begin{verbatim}
bits2 System.rounding_mode "IEEE 754 rounding mode";
\end{verbatim}
The name ``\texttt{System.rounding\_mode}'' is not special; it's the
string literal that makes this name refer to the hardware.
\item
You can't usefully assign to the rounding modes or read them---all you
can usefully do is use them in floating-point operations that require
rounding modes.
(This one might not be too hard to fix.)
\end{itemize}

\subsection{Widths of computation}

\begin{itemize}
\item
The back ends support computation only at the width of the target
machine.
Variables and operators must use this width.
(This one will be fixed soon.)
\end{itemize}

\subsection{Link-time constants}
\begin{itemize}
\item
The only link-time constants supported are those of the form ``symbol
+ offset.''
We can fix this bug relatively promptly if there is a need.
\end{itemize}



\subsection{Missing operations}

\begin{itemize}
\item
Lots of operations are missing, including
carry, borrow, extended multiplies, and anything else that might be
needed for large arithmetic.
\item
There's no automatic way to find out what operations are supported.
(But soon, the machine environment!)
\end{itemize}

\subsection{Features missing from the language}

\begin{itemize}
\item
The \texttt{also unwinds to} annotation does not work and may cause
things to fail.
\item
The \texttt{also returns to} annotation does not work and may cause
things to fail.
\end{itemize}


\subsection{Features missing from the run-time interface}
\begin{itemize}
\item
Anything to do with threads is probably not there, or if there, is a
little wacky.
\item
Unwinding the stack is not yet supported.
\end{itemize}

\subsection{Poor performance}

The compiler generates very bad code.
The implementation of the \texttt{switch} statement is especially
na\"\i ve.


\section{Notes for release 20040514}

We now have a very simple peephole optimizer.


\section{Notes for release 20040504}

Here's what you can do with a narrow variable or memory location:
\begin{itemize}
\item
Assign it a compile-time constant expression
\item
Assign it the contents of another variable or memory location of the
same width
\item
Assign it \texttt{\char`\%lobits} of an expression whose width is the
natural width of the target machine
\end{itemize}


\section{Notes for release 20040501}

We've added limited support for 64-bit variables on 32-bit platforms.
You can perform the following operations on a 64-bit {\PAL} variable:
\begin{itemize}
\item Assign it from a 64-bit literal
\item   Load it from memory
\item    Store it to memory
\item   Pass it to a C function
\item   Receive it from a C function
\end{itemize}
For the moment, actual operations on 64-bit values have to be done by
calling out to a C~compiler.
Here, for example, is a {\PAL}
procedure that tests 64-bit add:
\begin{verbatim} 
  target byteorder little;
  
  import ladd, printf;
  export main;
  
  foreign "C" main(bits32 argc, "address" bits32 argv) {
    bits64 n, m;
    n = 0x80000000::bits64;
    m = 0x80000000::bits64;
    bits64 sum;
    sum = foreign "C" ladd(n, m);
    foreign "C" printf(format, n, m, sum);
    foreign "C" return(0);
  }

  section "data" {
    format: bits8[] "sum of %016llx\nand    %016llx\nis     %016llx\n\0";
  }
\end{verbatim}
Here the imported function \texttt{ladd} is defined in~C as follows:
\begin{verbatim} 
  long long ladd(long long x, long long y) {
    return x + y;
  }
\end{verbatim}
The 64-bit variables are not actually stored in 32-bit register pairs,
but that optimization will probably come in a future release.


\section{Notes for release 20040427}

Quick~{\PAL} has a new front end.
Error messages should be more informative, and the front end now
detects some bugs that were undetected by the old front end.

In case of unforeseen difficulty, it is possible to use the old front
end by
\begin{verbatim}
  qc-- -e 'Compile.Old.use()' ...
\end{verbatim}
but this option is deprecated and will disappear in a future release.
Any problems with the new front end should be reported to
\texttt{bugs@cminusminus.org}. 


\section{Notes for release 20040405}
\begin{itemize}
\item
Unicode string literals are not implemented

\item Integer widening has been added, but not enabled.
In order to use widening, you must swap the order of the
``widen'' and ``placevars'' phases.  Look for the following lines in
``luacompile.nw'':
\begin{verbatim}
      Opt.verb(backend.placevars)  or Stages.null
    , Opt.verb(backend.widen)      or Stages.null
\end{verbatim}
The backend ``x86w'' shows how to construct a backend that does
integer widen properly for x86.

\end{itemize}



\section{Notes for release 20031021}
\begin{itemize}
\item
The compiler now generates runtime data. The native runtime system now
lives in the runtime directory.
\item
The interpreter interface has been changed slightly to be more
consistent with the native runtime interface. However, there are still
a few differences between the two interfaces. Interpreter functions
take pointers to activation structures where the native runtime takes
pointers to activation structures.
\item
The default register allocator has been changed to DLS. If you need
the graph coloring allocator, you can change it on the command line,
e.g.,
\begin{verbatim}
qc-- backend=Backend.x86 backend.ralloc=color ...
\end{verbatim}


\end{itemize}

\section{Notes for more recent releases}

\begin{itemize}
\item
If two different {\PAL} units have different global-variable
declarations, the error manifests as a failure to link.
\end{itemize}


\section{Notes for release 20030711}


This is the first major release that includes a native-code back end.
The release is believed consistent with the pre-2.0 manual (CVS
revision~1.75) with exceptions noted below.
\begin{itemize}
\item
The front end does not implement the \texttt{switch} statement.
\item
The \texttt{import} statement requires a type, although this
requirement is not documented in the manual.
\item
The native x86 back end successfully compiles the test suite
for the \texttt{lcc} C~compiler.
The back end has significant limitations:
\begin{itemize}
\item
Integer variables should be 8, 16, or 32~bits.
Integer and logical operations should be on 32-bit values only.
The back end supports the operations you would find in a C~compiler,
but we can try to add others on request.
\item
Floating-point operations may be at 32 or 64~bits.
80-bit floating-point is possible but not tested.
The back end supports the operations you would find in a C~compiler,
but we can try to add others on request.
\item 
Mixed integer and floating-point operations have not been tested
thoroughly. 
\item
The back end does not support multiprecision arithmetic or overflow
detection. 
\end{itemize}
\item
The release also includes back ends for Alpha and MIPS R3000, but these are
incomplete and have been used only to test implementations of calling
conventions. 
\item
There is as yet no run-time system to go with the native-code
compiler.
If you need a run-time system, you must use the interpreter.
\item
 Code quality is poor; as a matter of policy, we are postponing
       work on optimization in order to bring you a run-time system.
\item
The default register allocator is the \texttt{dls} register allocator.
This allocator is dramatically faster than our implementation of the
graph-coloring register allocator (\texttt{color}).
If you want, you may invoke the \texttt{color} allocator on the
command line, as in the following example:
\begin{verbatim} 
  qc-- Backend.x86.ralloc=Ralloc.color hello.c--
\end{verbatim}
\end{itemize}

\end{document}
